---
title: "Honey Bees Colonies in the US"
author: "dweorh"
date: "2023-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Business requirement

A software company is interested in writing a software for beekeepers to help them manage colonies and apiaries.

The company needs to know how big and stable the market is.

## Data source

For the analysis survey data collected by [United States Department of Agriculture](https://www.nass.usda.gov/) between 2012 and 2022 will be used.

## Project setup

data file: USDA_NASS_honey_bees_2012_2022_survey.csv

```{r libraries}
# install.packages('tidyverse')
library(tidyverse)
# install.packages("dplyr")
library("dplyr")
# install.packages("janitor")
library("janitor")
# install.packages("scales")
library("scales")
# install.packages("RColorBrewer")
library("RColorBrewer")

# color pallete 
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

expected_items <- c('ADDED & REPLACED',
                    'INVENTORY, MAX',
                    'LOSS, COLONY COLLAPSE DISORDER',
                    'LOSS, DEADOUT')

expected_periods <- c(1:4)
```

## Data cleanup

```{r data-cleanup}
survey <- read_csv('USDA_NASS_honey_bees_2012_2022_survey.csv')

# select only needed columns from the set
# normalize names
# remove redundant data
# convert text values to less bug prone and saving resources integer values
survey_reduced <- survey %>%
  clean_names() %>%
  select(year, period, state, data_item, value) %>%
  mutate(data_item = str_replace(data_item, 'HONEY, BEE COLONIES - ', '')) %>% 
  mutate(data_item = str_replace(data_item, ', MEASURED IN COLONIES', '')) %>%
  mutate(period = str_replace(period, 'JAN THRU MAR', '1')) %>% 
  mutate(period = str_replace(period, 'APR THRU JUN', '2')) %>% 
  mutate(period = str_replace(period, 'JUL THRU SEP', '3')) %>%
  mutate(period = str_replace(period, 'OCT THRU DEC', '4')) %>% 
  mutate(period = as.numeric(period))

# for further use save cleanup data
write_csv(survey_reduced, 'USDA_NASS_honey_bees_2012_2022_survey_reduced.csv')

# remove unused variables and release memory
rm(survey)
gc()

# check if there is correct number of periods (4) for every year and every state
survery_missing_periods <- survey_reduced %>%
  select(year,state,period) %>%
  group_by(year,state,period) %>%
  distinct() %>%
  group_by(year,state) %>%
  summarize(periods=n()) %>%
  filter(periods < length(expected_periods))

# if there are any missing periods for any year,
# we will exclude such a year from the further calculations
survey_missing_periods_years <- as.vector(t(survery_missing_periods %>% select(year) %>% distinct()))

```
Check if we have equal number of rows for each 'Data Item' type.
```{r data item qty check}
survey_reduced %>% group_by(data_item) %>% summarize(total_count=n())
```
Many 'LOSS, COLONY COLLAPSE DISORDER' data is missing.

Chart with totals, should show us if that's an issue.

As this is a survey, in some states in some years 'losses' might be merged into one category.

```{r country overview data}
survey_totals_per_state <- survey_reduced %>% 
  filter(!year %in% survey_missing_periods_years) %>% 
  filter(data_item == 'INVENTORY, MAX') %>%
  group_by(year,state) %>%
  summarize(total=sum(value))

survey_totals <- survey_reduced %>% group_by(year,data_item) %>% summarize(total = sum(value))

# a function to calculate totals of certain type
add_loss_total <- function(type = 'add') {
  data <- pick(data_item, total)
  # data <- cur_data()
  if (type == 'add') {
    res <- sum(data[data$data_item == 'ADDED & REPLACED',]$total)
  } else if (type == 'loss') {
    sub_data <- data[data$data_item %in% c('LOSS, DEADOUT', 'LOSS, COLONY COLLAPSE DISORDER'),] 
    res <- sum(sub_data$total)
  } else if (type == 'inventory') {
    res <- sum(data[data$data_item == 'INVENTORY, MAX',]$total)
  }
  return(res)
}

survey_totals_add_loss <- survey_totals %>%
  filter(!year %in% survey_missing_periods_years) %>% 
  group_by(year) %>%
  summarize(
    loss = add_loss_total('loss') / 1000,
    add = add_loss_total('add') / 1000,
    inventory = add_loss_total('inventory') / 1000
  )

totals_pivot <- pivot_longer(survey_totals_add_loss, cols = c(loss,add,inventory))

survey_last_valid_year = max(survey_totals_add_loss$year)

last_year_data <- totals_pivot %>% filter(year == survey_last_valid_year)
last_year_data_states <- survey_totals_per_state %>% filter(year == survey_last_valid_year)

```

A chart to present a general country level overview of 'add', 'loss', and 'inventory'.

```{r country overview chart}
labels <- list(add = "Added", loss = "Lost", inventory = "Total")
colors <- list(add = "green", loss = "red", inventory = "blue")

chart_title <- paste0("Honey Bees Colonies in The US (",min(survey_reduced['year']),"-",survey_last_valid_year,")")
chart_subtitle <- paste0("Missing years: ", paste(survey_missing_periods_years, collapse=", "))

ggplot(data=totals_pivot) + 
  geom_smooth(mapping=aes(x=year, y=value, color=name)) +
  labs(title=chart_title,
       subtitle = chart_subtitle,
    caption="Data collected by https://www.nass.usda.gov/",
     x='Year', y='No. of Colonies (in K)') +
  expand_limits(y=0) +
  geom_text(
    data = last_year_data,
    aes(label = labels[name], x = year + 0.2, y = ifelse(last_year_data$name=='add', value*0.9, value), color = name),
    parse = TRUE
  ) +
  scale_color_manual(name="", labels = labels, values = colors) +
  theme_bw() + 
  theme(legend.position = "none")
```
On the country level the total inventory stays more or less stable in 10y period.

The same with adds and losses.

## Detailed charts

To make charts more readable, states were divided, arbitrarily, into 4 categories:

* up to 50,000 colonies
* 50,000 - 100,000 colonies
* 100,000 - 500,000 colonies
* 500,000 colonies and more


```{r size categories}
v_small_states = as.vector(
  t(
    survey_totals_per_state %>%
      filter(year == 2021 & total < 50000) %>%
        ungroup() %>% select(state)
    )
  )
small_states = as.vector(
  t(
    survey_totals_per_state %>%
      filter(year == 2021 & total >= 50000 & total < 100000) %>%
      ungroup() %>% select(state)
    )
  )
mid_states = as.vector(
  t(
    survey_totals_per_state %>%
      filter(year == 2021 & total >= 100000 & total < 500000) %>%
      ungroup() %>% select(state)
    )
  )
big_states = as.vector(
  t(
    survey_totals_per_state %>%
      filter(year == 2021 & total >= 500000) %>%
      ungroup() %>% select(state)
    )
  )
```

```{r state charts}
chart_big_title <- paste0("H. Bees Col., 500,000+ colonies (",min(survey_reduced['year']),"-",survey_last_valid_year,")")

plot_data <- survey_totals_per_state %>% filter(state %in% big_states)
ggplot(data=plot_data) + 
  geom_smooth(mapping=aes(x=year, y=total, color=state)) +
  labs(title=chart_big_title,
       subtitle = chart_subtitle,
       caption="Data collected by https://www.nass.usda.gov/",
       x='Year', y='No. of Colonies') +
  expand_limits(y=0) +
  theme_bw() +
  scale_color_manual(name="State",
                     values = col_vector,
                     labels = paste(plot_data$state,
                                    scales::comma(plot_data[plot_data$year == survey_last_valid_year,]$total/1000),
                                    'K')
                     ) +
  scale_y_continuous(labels = scales::comma)

chart_mid_title <- paste0("H. Bees Col., 100,000 - 500,000 colonies (",min(survey_reduced['year']),"-",survey_last_valid_year,")")

plot_data <- survey_totals_per_state %>% filter(state %in% mid_states)
ggplot(data=plot_data) + 
  geom_smooth(mapping=aes(x=year, y=total, color=state)) +
  labs(title=chart_mid_title,
       subtitle = chart_subtitle,
       caption="Data collected by https://www.nass.usda.gov/",
       x='Year', y='No. of Colonies') +
  expand_limits(y=0) +
  theme_bw() +
  scale_color_manual(name="State",
                     values = col_vector,
                     labels = paste(plot_data$state,
                                    scales::comma(plot_data[plot_data$year == survey_last_valid_year,]$total/1000),
                                    'K')
  ) +
  scale_y_continuous(labels = scales::comma)

chart_small_title <- paste0("H. Bees Col., 50,000 - 100,000 colonies (",min(survey_reduced['year']),"-",survey_last_valid_year,")")

plot_data <- survey_totals_per_state %>% filter(state %in% small_states)
ggplot(data=plot_data) + 
  geom_smooth(mapping=aes(x=year, y=total, color=state)) +
  labs(title=chart_small_title,
       subtitle = chart_subtitle,
       caption="Data collected by https://www.nass.usda.gov/",
       x='Year', y='No. of Colonies') +
  expand_limits(y=0) +
  theme_bw() +
  scale_color_manual(name="State",
                     values = col_vector,
                     labels = paste(plot_data$state,
                                    scales::comma(plot_data[plot_data$year == survey_last_valid_year,]$total/1000),
                                    'K')
  ) +
  scale_y_continuous(labels = scales::comma)


chart_v_small_title <- paste0("H. Bees Col., 50,000 < colonies (",min(survey_reduced['year']),"-",survey_last_valid_year,")")

plot_data <- survey_totals_per_state %>% filter(state %in% v_small_states)
ggplot(data=plot_data) + 
  geom_smooth(mapping=aes(x=year, y=total, color=state)) +
  labs(title=chart_v_small_title,
       subtitle = chart_subtitle,
       caption="Data collected by https://www.nass.usda.gov/",
       x='Year', y='No. of Colonies') +
  expand_limits(y=0) +
  theme_bw() +
  scale_color_manual(name="State",
                     values = col_vector,
                     labels = paste(plot_data$state,
                                    scales::comma(plot_data[plot_data$year == survey_last_valid_year,]$total/1000),
                                    'K')
  ) +
  scale_y_continuous(labels = scales::comma)
```

## Conclusions
1. Honey Bees market looks stable across 10y, according to surveys.
2. The biggest beekeeping market is in California with ca. 30% of total US H. Bees colonies.
3. That market neither grows nor shrinks, so competition can be hard.
4. Further research about already existing software for beekeepers is needed. 
